Typing a linear \picalc

### 2020-04-12 paper/paper.tex
Introduce substitution

@@

Data type definitions have double lines and index-free synonyms as rule names, for ease of reference.
Otherwise, we use the constructor name as the name of a typing rule.
We omit universe levels for brevity.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructors can be shared across types and disambiguated by context.

@@

  \begin{remark}
    The constructor $\constr{zero}$ is the only place in the typing system that uses the algebra on multiplicities.
  \end{remark}

  \todo{example}

@@

    \inferrule
        {t : \Type \\
          x \; y \; z : \Carrier_{idx}^2\\
          \opsquared{x}{y}{z}}
    {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

    \inferrule
        {t \; t' \; : \Type \\
          x : \Carrier_{idx}^2 \\
          x' : \Carrier_{idx'}^2 \\\\
      loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
    {\constr{suc} \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
  \end{mathpar}
  \caption{Variable references as proofs on the typing context and input and output multiplicity contexts.}
  \label{fig:variable-references}
  \end{figure}
\end{nidefinition}

\begin{lemma}
  The typing context $\gamma$ at index $i$ is type $t$.
\end{lemma}
\begin{lemma}
  The input annotation in context $\Gamma$ at index $i$ is the result of aggregating (as per the algebra) $y$ and the output annotation in context $\Delta$ at index $i$.
\end{lemma}
\begin{lemma}
  For every index $j$ other than $i$, the input annotation in context $\Gamma$ at index $i$ is the output annotation in $\Delta$ at index $i$.
\end{lemma}
\begin{corollary}
  Given the witness $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$, $\Gamma \equiv \Delta$ iff $y \equiv \zero$.
\end{corollary}
\proof{All proofs are by structural induction on \textsc{VarRef}.}

@@

@@

\todo{op but for contexts}
\begin{nitheorem}
  Let $\opctx{\Gamma_l}{\Delta}{\Xi_l}$ and $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  If $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$ then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}

@@

\subsection{Subject Reduction}

\begin{nitheorem}
  If $P \reduce{c} Q$ and $\types{\gamma}{\Gamma'}{P}{\Xi}$ where $\opctx{\Gamma'}{\ell_{\#} @ c}{\Gamma}$,
  then $\types{\gamma}{\Gamma}{Q}{\Xi}$
\end{nitheorem}

>>>>

% Diagrams
\usepackage{tikz}

@@

Data type definitions have double lines and index-free synonyms as rule names, for ease of reference.
Otherwise, we use the constructor name as the name of a typing rule.
We omit universe levels for brevity.
Implicit arguments are mentioned by type definitions but omitted in constructors.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructors can be shared across types and disambiguated by context.

@@

@@

    \inferrule
    {\opsquared{x}{y}{z}}
    {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

    \inferrule
    {loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
    {\constr{suc} \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
  \end{mathpar}
  \caption{Variable references as proofs on the typing context and input and output multiplicity contexts.}
  \label{fig:variable-references}
  \end{figure}
\end{nidefinition}

\begin{remark}
  We use $\field{\cdot-unique}$ to relieve the user from the proof burden $\opsquared{x}{y}{z}$ in the base case $\constr{zero}$.
  This is the only place in the typing system that uses the algebra on multiplicities.
\end{remark}

\begin{example}
\todo{example}
\end{example}

\begin{lemma}
  The typing context $\gamma$ at index $i$ is type $t$.
\end{lemma}
\begin{lemma}
  The input annotation in context $\Gamma$ at index $i$ is the result of aggregating (as per the algebra) $y$ and the output annotation in context $\Delta$ at index $i$.
\end{lemma}
\begin{lemma}
  For every index $j$ other than $i$, the input annotation in context $\Gamma$ at index $i$ is the output annotation in $\Delta$ at index $i$.
\end{lemma}
\begin{lemma}
  \todo{Add feedback and feedfront}
\end{lemma}
\begin{corollary}
  Given the witness $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$, $\Gamma \equiv \Delta$ iff $y \equiv \zero$.
\end{corollary}
\proof{All proofs are by structural induction on \textsc{VarRef}.}

@@

Subject reduction (aka type preservation) relates the operational semantics and the typing system of a language.
In its usual form, it states that if $P$ reduces to $Q$ and $P$ is well-typed under context $\Gamma$, then $Q$ is well-typed under $\Gamma$ too.
In the linear \picalc{} however, reduction \emph{is} meant to transform the context: each time communication occurs over a channel, that channel is exhausted.
In our case, and because we generalize linearity to a partial monoid, the channel over which the communication occurs will loose a multiplicity on both its input and output capabilities.
In \S \ref{semantics}, we made sure to lift the index of this channel to the type level.
If the channel is internal to the process, the context must not be modified; if it is external on index $i$, the input usage context at index $i$ must have $\ell_\#$ subtracted.

\begin{nitheorem}
  Let $P \reduce{c} Q$.
  If $c$ is $\constr{internal}$, then let $\Delta \equiv \epsilon$;
  if $c$ is $\constr{external}\; i$, then let $\Delta \equiv \ell_\# @ i$.
  Let $\Gamma'$ and $\Gamma$ be usage contexts only differing in $\Delta$, so that $\opctx{\Gamma'}{\Delta}{\Gamma}$.
  If $\types{\gamma}{\Gamma'}{P}{\Xi}$ then $\types{\gamma}{\Gamma}{Q}{\Xi}$
\end{nitheorem}

\begin{proof}
\end{proof}

@@

\begin{nitheorem}
  Let $\opctx{\Gamma_l}{\Delta}{\Xi_l}$ and $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  If $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$ then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}

@@

When function abstraction and application come together in the \lambdacalc{} with de Bruijn indices, every reference to the innermost variable in the body of the abstraction gets replaced with the argument of the application.
In the \picalc{} however substitution takes another form: every reference to the innermost variable in the receiving process gets replaced with the reference to the data that is being sent over the channel.
In a resource-aware \picalc{}, that means that instead of exhausting all $m$ multiplicities of the innermost variable, we leave those unaltered and exhaust them from the element that is being sent.

\begin{nitheorem}
  Let $P$ be a process where the innermost variable has all its $m$ multiplities exhausted, such that $\types{\gamma , t}{\Gamma , m}{P}{\Psi , \ell_{\o}}$.
  Let $j$ be the index of the data that is being received.
  This data has at least $m$ multiplicities, such that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the innermost variable by $j$.
  As a result, we will exhaust no multiplicities on the innermost variable and exhaust them in $j$ instead, such that $\types{\gamma , t}{\Gamma , m}{\subst{P}{j}{\constr{zero}}}{\Xi , m}$.
\end{nitheorem}

\begin{proof}
  By induction on the typing derivation on $P$.
  To be able to proceed under scope restriction, we need to generalize the theorem so that the variable that is being substituted need not be $\constr{zero}$, but any variable $i$.
  We introduce such generalization in Theorem \ref{th:subst-generalization}.
  \todo{explain arguments}
\end{proof}

\begin{nitheorem}
  \label{th:subst-generalization}

  We introduce an intermediate usage context $\Psi_m$.
  This intermediate context is the result of exhausting some $\Delta$ from the input context $\Gamma$, such that $\opctx{\Gamma}{\Delta}{\Psi_m}$.
  From $\Psi_m$, exhausting $m$ at index $i$ yields $\Psi_l$; exhausting it at index $j$ yields $\Psi_j$, such that $\contains{\gamma}{\Psi_m}{i}{t}{m}{\Psi_l}$ and $\contains{\gamma}{\Psi_m}{j}{t}{m}{\Psi_r}$.
  The input derivation on $P$ takes context $\Gamma$ and yields context $\Psi_l$, such that $\types{\gamma}{\Gamma}{P}{\Psi_l}$.
  The output derivation on $\subst{P}{j}{i}$ takes the same context $\Gamma$ but yields back $\Psi_r$, such that $\types{\gamma}{\Gamma}{\subst{P}{j}{i}}{\Psi_r}$.

  \begin{remark}
    \label{re:delta-zero}
    It is of uttermost importance that the variable that is being substituted has \emph{all} of its multiplicities substituted.
    To enforce this, we add an additional premise: that in going from $\Gamma$ to the intermediate context $\Psi_m$ we must exhaust no multiplicities at index $i$ -- so that $m$ indeed captures the totality of the multiplicities that are being exhausted between $\Gamma$ and $\Psi_l$.
    We capture this requirement by enforcing $\Delta_i \equiv \ell_{\o}$.
  \end{remark}
\end{nitheorem}

\begin{figure}[ht]
  \begin{subfigure}{.5\textwidth}
  \centering
  \begin{tikzpicture}
    \node (gamma) at (0,2) {$\Gamma$};
    \node (psi-m) at (4,2) {$\Psi_m$};
    \node (psi-l) at (4,4) {$\Psi_l$};
    \node (psi-r) at (4,0) {$\Psi_r$};
    \draw[->]                (gamma) -- node[above,sloped] {$\Delta$} node[below] {$\Delta_i = \ell_{\o}$} (psi-m);
    \draw[->]                (gamma) -- node[above,sloped] {$\vdash P$}               (psi-l);
    \draw[->,densely dotted] (gamma) -- node[below,sloped] {$\vdash \subst{P}{j}{i}$} (psi-r);
    \draw[->]                (psi-m) -- node[right]        {$m@i$}                    (psi-l);
    \draw[->]                (psi-m) -- node[right]        {$m@j$}                    (psi-r);
  \end{tikzpicture}
  \caption{Diagrammatic representation of the generalized substitution theorem.}
  \label{fig:subst}
  \end{subfigure}
  % Do not leave empty line
  \begin{subfigure}{.5\textwidth}
  \centering
  \begin{tikzpicture}
    \node (gamma) at (0,2) {$\Gamma$};
    \node (xi-l)  at (0,4) {$\Xi_l$};
    \node (xi-r)  at (0,0) {$\Xi_r$};
    \node (psi-m) at (4,2) {$\Psi_m$};
    \node (psi-l) at (4,4) {$\Psi_l$};
    \node (psi-r) at (4,0) {$\Psi_r$};
    \draw[->]                (gamma) -- node[above,sloped] {$\Delta$} node[below] {$\Delta_i = \ell_{\o}$} (psi-m);
    \draw[->]                (gamma) -- node[left]         {$\ell_i@i$}               (xi-l);
    \draw[->,densely dotted] (gamma) -- node[left]         {$\ell_i@j$}               (xi-r);
    \draw[->]                (xi-l)  -- node[above]        {$\vdash P$}               (psi-l);
    \draw[->,densely dotted] (xi-r)  -- node[above]        {$\vdash \subst{P}{j}{i}$} (psi-r);
    \draw[->]                (psi-m) -- node[right]        {$m@i$}                    (psi-l);
    \draw[->]                (psi-m) -- node[right]        {$m@j$}                    (psi-r);
  \end{tikzpicture}
  \caption{
    Diagrammatic representation of the generalized substitution for process input.
    Thanks to the additional premise in Remark \ref{re:delta-zero}, we know that $m$ is at least capable of $\ell_i$, or $\opsquared{m}{\delta}{\ell_i}$.
  }
  \label{fig:subst-input}
  \end{subfigure}
\end{figure}

\begin{proof}
  \todo{BIG TODO}
\end{proof}
